---
title: 'Clean outbreaks data'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to clean and standardize case data?
- How to convert raw dataset into a `linelist` object?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Explain how clean, curate, and standardize case data using `{cleanepi}` package
- Demonstrate how to covert case data to a `linelist` object 

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction
In the process of analyzing outbreak data, it's essential to ensure that the dataset is clean, curated, and standardized to facilitate accurate and reproducible analysis. To achieve this, we will utilize the [cleanepi](https://epiverse-trace.github.io/cleanepi/) package. For demonstration purposes, we'll work with a simulated dataset of Ebola cases.

The first step is to import the dataset following the guidelines outlined in the [Read case data](../episodes/read-cases.Rmd) episode. This involves loading the dataset into our environment and view its structure and content. 

```{r}
requireNamespace("rio")
sim_ebola_data <- rio::import(file.path("data", "simulated_ebola.csv",
                                        fsep = "/"))
utils::head(sim_ebola_data, 5)
```

##  Quick inspection
Quick exploration and inspection of the dataset are crucial before diving into any  analysis tasks. The `{cleanepi}` package simplifies this process with the `scan_data()` function. Let's take a look at how you can use it:

```{r}
requireNamespace("cleanepi")
cleanepi::scan_data(sim_ebola_data)
```


The results provides a summary of each column, including column names, data types, number of missing values, and summary statistics for numerical columns.
