---
title: 'Input delay data'
teaching: 10
exercises: 2
editor_options: 
  chunk_output_type: inline
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to connect reused delays with my existing analysis pipeline?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Use distribution functions to continuous and discrete distributions stored as `<epidist>` objects.
- Convert a continuous to a discrete distribution with `{epiparameter}`.
- Connect `{epiparameter}` outputs with `{EpiNow2}` inputs.

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites

This episode requires you to be familiar with:

**Data science** : Basic programming with R.

**Epidemic theory** : Epidemiological parameters. Time periods.

:::::::::::::::::::::::::::::::::

## Introduction

Now, we have an epidemiological parameter we can reuse! We can replace the **summary statistics** numbers we plug into `EpiNow2::dist_spec()`.

Let's assign this `<epidist>` class object to the `covid_serialint` object.

```{r}
covid_serialint <-
  epiparameter::epidist_db(
    disease = "covid",
    epi_dist = "serial",
    author = "Nishiura",
    single_epidist = TRUE
  )

covid_serialint
```

<!-- to activate for EpiNow2@dist-interfase

But still, we need to extract them as usable numbers. We use `epiparameter::get_parameters()` for this:

```{r}
covid_serialint_parameters <- epiparameter::get_parameters(covid_serialint)

covid_serialint_parameters
```

This gets a vector of class `<numeric>` useful as input for any other package! 

:::::::::::::::::: callout

If we write the `[]` next to the last object create like in `covid_serialint_parameters[]`, within `[]` we can use the 
Tab key <kbd>↹</kbd> 
to use the [code completion feature](https://support.posit.co/hc/en-us/articles/205273297-Code-Completion-in-the-RStudio-IDE) and have a quick access to `covid_serialint_parameters["meanlog"]` and `covid_serialint_parameters["sdlog"]`. We invite you to try this out!

This only seems to work in code chunks and R console!

::::::::::::::::::

-->

But, how can we plug them into `{EpiNow2}`? In this lesson we are going to review the distributions functions that `{epiparameter}` provides to get the value we need for this and any other package downstream in the pipeline!

## Continuous distributions

The following output has four entries with different content in the **probability distribution** (`prob_distribution`) column:

```{r}
distribution <-
  epiparameter::epidist_db(
    disease = "covid",
    epi_dist = "serial"
  )

distribution %>%
  list_distributions()
```

Entries with a missing value (`<NA>`) in the `prob_distribution` column are *non-parameterised* entries. They have summary statistics but no probability distribution. Compare these two outputs:

```{r,eval=FALSE}
distribution %>%
  # pluck the first entry in the object class <list>
  pluck(1) %>%
  # check if <epidist> object have distribution parameters
  is_parameterised()
```

::::::::::::::::: spoiler

### Parameterised entries have an Inference method

As detailed in `?is_parameterised`, a parameterised distribution is the entry that has a probability distribution associated with it provided by an `inference_method` as shown in `metadata`:

```{r,eval=FALSE}
distribution[[1]]$metadata$inference_method
distribution[[2]]$metadata$inference_method
distribution[[4]]$metadata$inference_method
```

:::::::::::::::::::::::::

In the `epiparameter::list_distributions()` output, we can also find different types of probability distributions (e.g., Log-normal, Weibull, Normal).

```{r}
distribution %>%
  list_distributions()
```

In `{epiparameter}`, you will mostly find **continuous** distributions like these. You can visualize any of them with the `plot()` function and access to: 

- the *Probability Density Function (PDF)* and 
- the *Cumulative Distribution Function (CDF)*.

```{r}
# get a single <epidist> object
distribution_single <-
  epiparameter::epidist_db(
    disease = "covid",
    epi_dist = "serial",
    single_epidist = TRUE
  )

# plot <epidist> object
plot(distribution_single)
```

With the `day_range` argument, you can change the length or number of days in the `x` axis. Explore how this looks like:

```{r,eval=FALSE}
# plot <epidist> object
plot(distribution_single, day_range = 0:20)
```

::::::::::::::::: discussion

### The distribution Zoo

Explore this shinyapp called **The Distribution Zoo**!

Follow these steps to reproduce the form of the COVID serial interval distribution from `{epiparameter}` (`covid_serialint` object):

1. Access to <https://ben18785.shinyapps.io/distribution-zoo/> shinyapp website,
2. Go to the left panel,
3. Keep the *Category of distribution*: `Continuous Univariate`,
4. Select a new *Type of distribution*: `Log-Normal`,
5. Move the **sliders**, i.e. the graphical control element that allows you to adjust a value by moving a handle along a horizontal track or bar to the `covid_serialint` parameters. 

Replicate these with the `distribution` object and all its list elements: `[[2]]`, `[[3]]`, and `[[4]]`. Explore how the shape of a distribution changes when its parameters change.

Share about:

- What other features of the website do you find helpful?

:::::::::::::::::::::::::

::::::::::::::::::::::::: instructor

For shinyapps like the [Distribution Zoo](https://ben18785.shinyapps.io/distribution-zoo/):

A slider, in the context of user interfaces and graphical user interfaces (GUIs), is a graphical control element that allows users to adjust a value by moving a handle along a track or bar. Conceptually, it provides a way to select a numeric value within a specified range by visually sliding or dragging a pointer (the handle) along a continuous axis.

:::::::::::::::::::::::::

## Distribution functions

In R, all the statistical distributions have functions to access the:

- Probability **Density** function (PDF),
- Cumulative Distribution function (**CDF**),
- **Quantile** function, and
- **Random** values from the given distribution.

:::::::::::: spoiler

### Functions for the Normal distribution

If you need it, read in detail about the [R probability functions for the normal distribution](https://sakai.unc.edu/access/content/group/3d1eb92e-7848-4f55-90c3-7c72a54e7e43/public/docs/lectures/lecture13.htm#probfunc), each of its definitions and identify in which part of a distribution they are located!

![The four probability functions for the normal distribution ([Jack Weiss, 2012](https://sakai.unc.edu/access/content/group/3d1eb92e-7848-4f55-90c3-7c72a54e7e43/public/docs/lectures/lecture13.htm#probfunc))](fig/fig5a-normaldistribution.png)

::::::::::::::::::::

If you look at `?stats::Distributions`, each type of distribution has a unique set of functions. However, `{epiparameter}` gives you the same four functions to access each of the values above for any `<epidist>` object you want! 

```{r,eval=FALSE}
# plot this to have a visual reference
plot(covid_serialint, day_range = 0:20)
```

```{r}
# the density value at quantile value of 10 (days)
density(covid_serialint, at = 10)

# the cumulative probability at quantile value of 10 (days)
cdf(covid_serialint, q = 10)

# the quantile value (day) at a cumulative probability of 60%
quantile(covid_serialint, p = 0.6)

# generate 10 random values (days) given
# the distribution family and its parameters
generate(covid_serialint, times = 10)
```

::::::::: instructor

Access to the reference documentation (Help files) for these functions is accessible with the three double-colon notation: `epiparameter:::`

- `?epiparameter:::density.epidist()`
- `?epiparameter:::cdf.epidist()`
- `?epiparameter:::quantile.epidist()`
- `?epiparameter:::generate.epidist()`

::::::::::::::::::

::::::::::::::::::::::::::::::::: challenge

### Window for contact tracing and the Serial interval

The **serial interval** is important in the optimization of contact tracing since it provides a time window for the containment of a disease spread ([Fine, 2003](https://academic.oup.com/aje/article/158/11/1039/162725)). Depending on the serial interval, we can evaluate the need to expand the number of days pre-onset to consider in the contact tracing to include more backwards contacts ([Davis et al., 2020](https://assets.publishing.service.gov.uk/media/61e9ab3f8fa8f50597fb3078/S0523_Oxford_-_Backwards_contact_tracing.pdf)).

With the COVID-19 serial interval (`covid_serialint`) calculate:

- How much more of the backward cases could be captured if the contact tracing method considered contacts up to 6 days pre-onset compared to 2 days pre-onset?

::::::::::::::::: hint

In Figure 5 from the [R probability functions for the normal distribution](https://sakai.unc.edu/access/content/group/3d1eb92e-7848-4f55-90c3-7c72a54e7e43/public/docs/lectures/lecture13.htm#probfunc), the shadowed section represents a cumulative probability of `0.997` for the quantile value at `x = 2`.

::::::::::::::::::::::

::::::::::::::::: solution

```{r,eval=FALSE}
plot(covid_serialint)
```

```{r,eval=TRUE}
cdf(covid_serialint, q = 2)
cdf(covid_serialint, q = 6)
```

Given the COVID-19 serial interval:

- A contact tracing method considering contacts up to 2 days pre-onset will capture around 11.1% of backward cases.

- If this period is extended to 6 days pre-onset, this could include 76.2% of backward contacts.

::::::::::::::::::::::::::

::::::::::::::::: solution

### What if

If we exchange the question between days and cumulative probability to: 

- When considering secondary cases, how many days following the symptom onset of primary cases can we expect 55% of symptom onset to occur?

```{r,eval=FALSE}
quantile(covid_serialint, p = 0.55)
```

An interpretation could be:

- The 55% percent of the symptom onset of secondary cases will happen after 4.2 days after the symptom onset of primary cases.

::::::::::::::::::::::::::


:::::::::::::::::::::::::::::::::::::::::::


## Discretize a continuous distribution

We are getting closer to the end! `EpiNow2::dist_spec()` still needs a maximum value (`max`). 

One way to do this is to get the quantile value for the distribution's 99.9th percentile or `0.999` cumulative probability. For this, we need access to the set of distribution functions for our `<epidist>` object.

We can use the set of distribution functions for a _continuous_ distribution (as above). However, these values will be _continuous_ numbers. We can **discretize** the continuous distribution stored in our `<epidist>` object to get discrete values from a continuous distribution.

When we `epiparameter::discretise()` the continuous distribution we get a **discrete**(-ized) distribution:

```{r}
covid_serialint_discrete <-
  epiparameter::discretise(covid_serialint)

covid_serialint_discrete
```

We identify this change in the `Distribution:` output line of the `<epidist>` object. Take a double check to this line:

```
Distribution: discrete lnorm
```

While for a **continuous** distribution, we plot the *Probability Density Function (PDF)*, for a **discrete** distribution, we plot the *Probability Mass Function (PMF)*:

```{r,eval=FALSE}
# continuous
plot(covid_serialint)

# discrete
plot(covid_serialint_discrete)
```

To finally get a `max` value, let's access the quantile value of the 99.9th percentile or `0.999` probability of the distribution with the `prob_dist$q` notation, similarly to how we access the `summary_stats` values.

```{r}
covid_serialint_discrete_max <-
  quantile(covid_serialint_discrete, p = 0.999)
```

::::::::::::::::::::::::::::::::: challenge

### Lenght of quarantine and Incubation period

The **incubation period** distribution is a useful delay to assess the length of active monitoring or quarantine ([Lauer et al., 2020](https://www.acpjournals.org/doi/10.7326/M20-0504)). Similarly, delays from symptom onset to recovery (or death) will determine the required duration of health-care and case isolation ([Cori et al., 2017](https://royalsocietypublishing.org/doi/10.1098/rstb.2016.0371)).

Calculate:

- Within what exact time frame do 99% of individuals who develop COVID-19 symptoms exhibit them after infection?

::::::::::::::::: hint

What delay distribution measures the time between infection and the onset of symptoms?

The probability function for `<epidist>` **discrete** distributions are the same that we used for the *continuous* ones!

```{r,eval=FALSE}
# plot to have a visual reference
plot(covid_serialint_discrete, day_range = 0:20)

# density value at quantile value 10 (day)
density(covid_serialint_discrete, at = 10)

# cumulative probability at quantile value 10 (day)
cdf(covid_serialint_discrete, q = 10)

# In what quantile value (days) do we have the 60% cumulative probability?
quantile(covid_serialint_discrete, p = 0.6)

# generate random values
generate(covid_serialint_discrete, times = 10)
```

::::::::::::::::::::::

::::::::::::::::: solution

```{r,eval=TRUE}
covid_incubation <-
  epiparameter::epidist_db(
    disease = "covid",
    epi_dist = "incubation",
    single_epidist = TRUE
  )

covid_incubation_discrete <- epiparameter::discretise(covid_incubation)

quantile(covid_incubation_discrete, p = 0.99)
```

99% of those who develop COVID-19 symptoms will do so within 16 days of infection.

Now, _Is this result expected in epidemiological terms?_

::::::::::::::::::::::::::

::::::::::::::::: solution

### how to create a distribution plot?

From a maximum value with `quantile()`, we can create a sequence of quantile values as a numeric vector and calculate `density()` values for each:

```{r,eval=TRUE}
# create a discrete distribution visualization
# from a maximum value from the distribution
quantile(covid_serialint_discrete, p = 0.999) %>%
  # generate quantile values
  # as a sequence for each natural number
  seq(1L, to = ., by = 1L) %>%
  # coerce numeric vector to data frame
  as_tibble_col(column_name = "quantile_values") %>%
  mutate(
    # calculate density values
    # for each quantile in the density function
    density_values =
      density(
        x = covid_serialint_discrete,
        at = quantile_values
      )
  ) %>%
  # create plot
  ggplot(
    aes(
      x = quantile_values,
      y = density_values
    )
  ) +
  geom_col()
```

**Remember:** In infections with pre-symptomatic transmission, serial intervals can have negative values ([Nishiura et al., 2020](https://www.ijidonline.com/article/S1201-9712(20)30119-3/fulltext)). When we use the _serial interval_ as an approximation of the _generation time_ we need to make this distribution with positive values only!

::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::


## Plug-in `{epiparameter}` to `{EpiNow2}`

Now we can plug everything into the `EpiNow2::dist_spec()` function!

- the **summary statistics** `mean` and `sd` of the distribution,
- a maximum value `max`,
- the `distribution` name.

```{r}
serial_interval_covid <-
  dist_spec(
    mean = covid_serialint$summary_stats$mean,
    sd = covid_serialint$summary_stats$sd,
    max = covid_serialint_discrete_max,
    distribution = "lognormal"
  )

serial_interval_covid
```

:::::::::: callout

### Warning

Using the serial interval instead of the generation time is an alternative that can propagate bias in your estimates, even more so in diseases with reported pre-symptomatic transmission. ([Chung Lau et al., 2021](https://academic.oup.com/jid/article/224/10/1664/6356465))

::::::::::::::::::

Let's replace the `generation_time` input we used for `EpiNow2::epinow()`.

```{r,eval=FALSE}
epinow_estimates <- epinow(
  # cases
  reported_cases = example_confirmed[1:60],
  # delays
  generation_time = generation_time_opts(serial_interval_covid),
  # computation
  stan = stan_opts(
    cores = 4, samples = 1000, chains = 3,
    control = list(adapt_delta = 0.99)
  )
)

base::plot(epinow_estimates)
```

::::::::::::::::::::::::::::::::: challenge

### Ebola's effective reproduction number

Download and read the [Ebola dataset](data/ebola_cases.csv):

- Reuse one epidemiological parameter to estimate the effective reproduction number for the Ebola dataset.
- Why did you choose that parameter?

::::::::::::::::: hint

To calculate the $R_t$, we need:

- data set with confirmed cases per day and
- one key delay distribution

Key functions we applied in this episode are:

- `epidist_db()`
- `list_distributions()`
- `discretise()`
- probability functions for continuous and discrete distributions 

::::::::::::::::::::::

::::::::::::::::: solution

```{r,eval=TRUE,echo=FALSE}
# read data from the tutorial repository R project
ebola_confirmed <-
  read_csv(file.path("data", "ebola_cases.csv"))
```

```{r,eval=FALSE,echo=TRUE}
# read data
# e.g.: if path to file is data/raw-data/ebola_cases.csv then:
ebola_confirmed <-
  read_csv(here::here("data", "raw-data", "ebola_cases.csv"))

# list distributions
epidist_db(disease = "ebola") %>%
  list_distributions()
```

```{r,message=FALSE,warning=FALSE,eval=TRUE}
# subset one distribution
ebola_serial <- epidist_db(
  disease = "ebola",
  epi_dist = "serial",
  single_epidist = TRUE
)

# adapt epiparameter to epinow2
ebola_serial_discrete <- discretise(ebola_serial)

ebola_serial_discrete_max <- quantile(ebola_serial_discrete, p = 0.999)

serial_interval_ebola <-
  dist_spec(
    mean = ebola_serial$summary_stats$mean,
    sd = ebola_serial$summary_stats$sd,
    max = ebola_serial_discrete_max,
    distribution = "gamma" # don't forget! it's a must!
  )

# run epinow
epinow_estimates <- epinow(
  # cases
  reported_cases = ebola_confirmed,
  # delays
  generation_time = generation_time_opts(serial_interval_ebola),
  # computation
  stan = stan_opts(
    cores = 4, samples = 1000, chains = 3,
    control = list(adapt_delta = 0.99)
  )
)

plot(epinow_estimates)

```

::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::  

<!--
## Concept map

update it from last epiparameter test 
-->

::::::::::::::::::::::::::::::::::::: keypoints 

- Use distribution functions to get summary statistics and informative parameters for public health interventions from `<epidist>` objects.
- Use `discretise()` to convert continuous to discrete delay distributions.

::::::::::::::::::::::::::::::::::::::::::::::::

