---
title: 'Add more delays'
teaching: 10
exercises: 2
editor_options: 
  chunk_output_type: inline
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to get summary statistics from `<epidist>` objects with only distribution parameters?
- When should delays be reused from a systematic review?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Reuse reporting delays from `{epiparameter}` as `{EpiNow2}` inputs.
- Convert distribution parameters to summary statistics with `{epiparameter}`

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites

This episode requires you to be familiar with:

**Data science** : Basic programming with R.

**Epidemic theory** : Epidemiological parameters. Time periods.

:::::::::::::::::::::::::::::::::

## Introduction

We have practised how to get epidemiological parameters from the literature and used them as input for other packages.

You will find complementary challenges and resources to continue your learning path here!

```{r,warning=FALSE,message=FALSE}
library(epiparameter)
library(EpiNow2)
library(tidyverse)
```

## Convert parameters to summary statistics

::::::::::::::::::::::::::::::::: challenge

Use the `influenza_england_1978_school` from the `{outbreaks}` package to calculate the effective reproduction number.

::::::::::::::::: hint

How to get the mean and standard deviation from a generation time with only distribution parameters but no summary statistics like `mean` or `sd` for `EpiNow2::dist_spec()`?

- Look at how to extract parameters from `{epiparameter}` vignette on [parameter extraction and conversion](https://epiverse-trace.github.io/epiparameter/articles/extract_convert.html)

::::::::::::::::::::::

::::::::::::::::: solution

<!-- https://github.com/epiverse-trace/epiparameter/issues/114 -->

```{r,eval=TRUE}
# What parameters are available for Influenza?
epidist_db(disease = "influenza") %>%
  list_distributions() %>%
  as_tibble() %>%
  count(epi_distribution)

influenza_generation <-
  epidist_db(
    disease = "influenza",
    epi_dist = "generation"
  )

influenza_generation_discrete <-
  discretise(influenza_generation)

# we have a problem
# the summary statistics do not have mean and sd
influenza_generation$summary_stats

# one solution is to
# get parameters and convert to summary statistics

# first,
# get parameters
influenza_generation_params <-
  get_parameters(influenza_generation)

# then,
# convert distribution parameters to summary statistics
influenza_converted <-
  convert_params_to_summary_stats(
    distribution = "weibull",
    shape = influenza_generation_params["shape"],
    scale = influenza_generation_params["scale"]
  )

influenza_converted

# however, dist_spec() only accepts lognormal or gamma distributions
# then, we can rely on getting the pmf
# this follows the same code as for the ggplot2 above!

influenza_generation_pmf <-
  # from a maximum value from the distribution
  quantile(influenza_generation_discrete, p = 0.999) %>%
  # generate quantile values
  # as a sequence for each natural number
  seq(1L, to = ., by = 1L) %>%
  # coerce numeric vector to data frame
  as_tibble_col(column_name = "quantile_values") %>%
  # calculate density values
  # for each quantile in the density function
  mutate(
    density_values = density(
      x = influenza_generation_discrete,
      at = quantile_values
    )
  )

influenza_generation_pmf

generation_time_influenza <-
  dist_spec(
    pmf = influenza_generation_pmf %>% pull(density_values)
  )

influenza_cleaned <-
  outbreaks::influenza_england_1978_school %>%
  select(date, confirm = in_bed)

epinow_estimates <- epinow(
  # cases
  reported_cases = influenza_cleaned,
  # delays
  generation_time = generation_time_opts(generation_time_influenza),
  # computation
  stan = stan_opts(
    cores = 4, samples = 1000, chains = 3,
    control = list(adapt_delta = 0.99)
  )
)

plot(epinow_estimates)
```

::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::

## When to reuse? When to estimate?

In the early stage of an outbreak, we can rely on reusing parameters for known pathogens to unknown ones, like for the [Disease X](https://en.wikipedia.org/wiki/Disease_X), a pathogen currently unknown to cause human disease and potentially cause a serious international epidemic ([WHO, 2018](https://www.who.int/activities/prioritizing-diseases-for-research-and-development-in-emergency-contexts)). 

But when data from **lines list paired with contact tracing** is available, we can estimate the key delay distributions that best fit our data. These will help us to inform, compare and update any previous estimate about questions like:

- How long should contacts be followed?
- What is the required duration of contact tracing?
- How long should cases be isolated to reduce transmission?

However, the methods to accurately estimate delays like the generation interval from contact tracing data involve adjusting for biases like **[censoring](../learners/reference.md#censoring)**, **[right truncation](../learners/reference.md#truncation)** and **[epidemic phase bias](../learners/reference.md#phasebias)**. [(Gostic et al., 2020)](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1008409)

We can identify what entries in the `{epiparameter}` library assessed for these biases in their methodology with the `method_assess` nested entry:

```{r}
covid_serialint$method_assess
```

::::::::::::::::: testimonial

### How to estimate delay distributions for Disease X?

Refer to this excellent tutorial on estimating the serial interval and incubation period of Disease X accounting for *censoring* using Bayesian inference with packages like `{rstan}` and `{coarseDataTools}`.

- Tutorial in English: <https://rpubs.com/tracelac/diseaseX> <!-- to request -->
- Tutorial en Español: <https://epiverse-trace.github.io/epimodelac/EnfermedadX.html>

**Then,** after you get your estimated values, you can manually create your own` <epidist>` class objects with `epiparameter::epidist()`! Take a look at its [reference guide on "Create an `<epidist>` object"](https://epiverse-trace.github.io/epiparameter/reference/epidist.html#ref-examples)!

:::::::::::::::::::::::::::::

:::::::::::::::::::::::::: spoiler

### How to assess for pre-symptomatic infection?

The lengths of the Serial interval and Incubation period determine the type of disease transmission. 

![The relationship between the incubation period and serial interval. From [Nishiura 2020](https://www.ijidonline.com/article/S1201-9712(20)30119-3/fulltext)](fig/incubation-period-serial-interval.jpg)

Estimating the proportion of pre-symptomatic infections, or the extent to which infectiousness precedes symptom onset will determine the effectiveness of contact tracing and the feasibility of controlling an outbreak ([Fraser et al., 2004](https://www.pnas.org/doi/full/10.1073/pnas.0307506101) and [Hellewell et al., 2020](https://www.thelancet.com/article/S2214-109X(20)30074-7/fulltext)).

<!-- The proportion of pre-symptomatic transmission contributes to the effective reproduction number and expected new cases. -->

![Parameter estimates. Plausible ranges for the key parameters R0 and θ (read the main text for sources) for four viral infections of public concern are shown as shaded regions. The size of the shaded area reflects the uncertainties in the parameter estimates. [Fraser et al., 2004](https://www.pnas.org/doi/10.1073/pnas.0307506101)](fig/reproduction-number-pre-symptomatic.png)

Meta-analysis on the proportion of pre-symptomatic and asymptomatic transmission in SARS-CoV-2 found limitations of the evidence given high heterogeneity and high risk of selection and information bias between studies ([Buitrago-Garcia et al., 2022](https://journals.plos.org/plosmedicine/article/figure?id=10.1371/journal.pmed.1003987.g003)). This is a call to action to improve the Outbreak Analytic pipelines to use and reuse in the early phase of an outbreak.

:::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::: challenge

### What type of transmission?

Compare the serial interval and incubation period of Influenza and MERS:

- What type of transmission may Influenza have?
- What type of transmission may MERS have?
- Do these results correlate with the available evidence?

::::::::::::::::: hint

For types of transmission, we refer to infections with symptomatic or pre-symptomatic transmission.

Key functions:

- `epidist_db()`
- `epidist$summary_stats$`

::::::::::::::::::::::

::::::::::::::::: solution

In this solution we use `purrr::pluck()` to extract elements within the `summary_stats` object which is of class `list`.

```{r,eval=TRUE}
# pre-symptomatic transmission
epidist_db(
  disease = "influenza",
  epi_dist = "incubation",
  single_epidist = TRUE
) %>%
  pluck("summary_stats") %>%
  pluck("mean")

epidist_db(
  disease = "influenza",
  epi_dist = "serial",
  single_epidist = TRUE
) %>%
  pluck("summary_stats") %>%
  pluck("mean")

# symptomatic transmission
epidist_db(
  disease = "mers",
  epi_dist = "incubation",
  single_epidist = TRUE
) %>%
  pluck("summary_stats") %>%
  pluck("median")

epidist_db(
  disease = "mers",
  epi_dist = "serial",
  single_epidist = TRUE
) %>%
  pluck("summary_stats") %>%
  pluck("mean")
```

```{r,eval=TRUE}
# pre-symptomatic transmission
epidist_db(
  disease = "covid",
  epi_dist = "incubation",
  author = "Stephen",
  single_epidist = TRUE
) %>%
  pluck("summary_stats") %>%
  pluck("mean")

epidist_db(
  disease = "covid",
  epi_dist = "serial",
  author = "Nishiura",
  single_epidist = TRUE
) %>%
  pluck("summary_stats") %>%
  pluck("mean")

# symptomatic transmission
epidist_db(
  disease = "ebola",
  epi_dist = "incubation",
  single_epidist = TRUE
) %>%
  pluck("summary_stats") %>%
  pluck("mean")

epidist_db(
  disease = "ebola",
  epi_dist = "serial",
  single_epidist = TRUE
) %>%
  pluck("summary_stats") %>%
  pluck("mean")
```

::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::


<!--
## Concept map

update it from last epiparameter test 
-->

::::::::::::::::::::::::::::::::::::: keypoints 

- Use `{epiparameter}` to get reporting delays required in transmissibility estimates. 
- Reuse known estimates for unknown disease in the early stage of an outbreak when no contact tracing data is available.

::::::::::::::::::::::::::::::::::::::::::::::::

