---
title: 'Aggregate and visualize'
teaching: 20
exercises: 10
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to aggregate case data? 
- How to visualize aggregated data?
- What is distribution of cases in time, place, gender, age?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Simulate synthetic outbreak data
- Convert linelist data to incidence 
- Create epidemic curves from incidence data
::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

In an analytic pipeline, exploratory data analysis (EDA) is an important step before formal modelling. EDA helps 
determine relationships between variables and summarize their main characteristics often by means of data visualization. 

This episode focuses on  EDA of outbreaks and epidemic data, and how to achieved that using a couples of handy `R` 
packages. A key observation in EDA of epidemic analysis is capturing the relationship between time and the number of 
reported cases, spanning various categories (confirmed, hospitalized, deaths, and recoveries), locations, and other 
demographic factors such as gender, age, etc.  


::::::::::::::::::: checklist

### The double-colon

The double-colon `::` in R let you call a specific function from a package without loading the entire package into the current environment. 

For example, `dplyr::filter(data, condition)` uses `filter()` from the `{dplyr}` package.

This help us remember package functions and avoid namespace conflicts.

:::::::::::::::::::

 
## Synthetic outbreak data

To illustrate the process of conducting EDA on outbreak data, we will generate a line list 
for a hypothetical Ebola outbreak utilizing the `{simulist}` package. This line list dataset offers individual-level 
information about the outbreak. For our simulation, we will assume that the dynamics of this outbreak are influenced by 
several factors: the contact distribution (average number of contacts for an infected case), distribution of contact 
intervals (time period between contacts), and the delay distributions of onset to hospitalization and onset to death. 
These latter distributions can be sourced from literature and are conveniently available in the `{epiparameter}` 
package, see the below code chunk.

```{r, warning=FALSE, message=FALSE}
# Load simulist and epiparameter packages
library("simulist")
library("epiparameter")

# Define contact distribution
contact_dist <- epiparameter::epidist(
  disease = "Ebola",
  epi_dist = "contact distribution",
  prob_distribution = "pois",
  prob_distribution_params = c(mean = 2)
)

# Define  distribution for interval between contact
cont_interval <- epiparameter::epidist(
  disease = "Ebola",
  epi_dist = "contact interval",
  prob_distribution = "gamma",
  prob_distribution_params = c(shape = 1, scale = 1)
)

# Define onset to hospitalized distribution
onset_to_hosp <- contact_dist <- epiparameter::epidist(
  disease = "Ebola",
  epi_dist = "onset to hospitalisatio",
  prob_distribution = "pois",
  prob_distribution_params = c(mean = 7)
)

# get onset to death from {epiparameter} database
onset_to_death <- epiparameter::epidist_db(
  disease = "Ebola",
  epi_dist = "onset to death",
  single_epidist = TRUE
)

# Define distribution for infectious period
infect_period <- epiparameter::epidist(
  disease = "Ebola",
  epi_dist = "Infectious period",
  prob_distribution = "gamma",
  prob_distribution_params = c(shape = 1, scale = 1)
)
```
Additionally, we assume that the outbreak started at the beginning of 2023, is highly contagious with a probability of 
infection of $80\%$, and its minimum and maximum sizes are 1000 and 10,000, respectively. Combining these assumptions with 
the mentioned distributions, the code chunk below generates a simulated line list:
```{r, warning=FALSE, message=FALSE}
# Set seed to 1 to  have the same results
base::set.seed(1)

# Generate simulation data using the defined distribution.
linelist <- simulist::sim_linelist(
  contact_dist,
  infect_period,
  prob_infect = 0.6,
  onset_to_hosp,
  onset_to_death,
  hosp_risk = 0.2,
  hosp_death_risk = 0.5,
  non_hosp_death_risk = 0.05,
  outbreak_start_date = as.Date("2023-01-01"),
  outbreak_size = c(1000, 10000),
  population_age = c(1, 90),
  case_type_probs = c(suspected = 0.2, probable = 0.1, confirmed = 0.7),
  config = simulist::create_config()
)

# View first few rows of the generated data
utils::head(linelist)
```
## Aggregating

Downstream analysis involves working with aggregated data rather than individual cases. This requires grouping linelist 
data in the form of incidence data. The [incidence2]((https://www.reconverse.org/incidence2/articles/incidence2.html){.external target="_blank"}) 
package offers an essential function, called `incidence`, for grouping case data, usually centered around dated events 
and/or other factors. The code chunk provided below demonstrates the creation of an `incidence2` object from the 
simulated  Ebola `linelist` data based on the  date of onset.

```{r, message=FALSE, warning=FALSE}
# load incidence2 package
library("incidence2")

# create incidence object by aggregating case data  based on the date of onset
dialy_incidence_data <- incidence2::incidence(
  linelist,
  date_index = "date_onset",
  interval = 1
)

# View the first incidence data for the first 5 days
utils::head(dialy_incidence_data, 5)
```
Furthermore, with the `{incidence2}` package, you can specify the desired interval and categorize cases by one or 
more factors. Below is a code snippet demonstrating weekly cases grouped by the date of onset and gender.

```{r}
# Grouping data by week
weekly_incidence_data <- incidence2::incidence(
  linelist,
  date_index = "date_onset",
  interval = 7,
  groups = c("sex", "case_type")
)

# View incidence data for the first 5 weeks
utils::head(weekly_incidence_data, 5)
```

::::::::::::::::::::::::::::::::::::: callout
## Notes 
When cases are grouped by different factors, it's possible that these groups may have different date ranges in the 
resulting `incidence2` object. The `incidence2` package provides a function called `complete_dates()` to ensure that an
 incidence object has the same range of dates for each group. By default, missing counts will be filled with 0.

```{r, message=FALSE, warning=FALSE}
# Create incidence object
dialy_incidence_data_2 <- incidence2::incidence(
  linelist,
  date_index = "date_onset",
  groups = "sex",
  interval = 1
)

# Complete missing dates in the incidence object
incidence2::complete_dates(
  x = dialy_incidence_data_2,
  expand = TRUE,
  fill = 0L, by = 1L,
  allow_POSIXct = FALSE
)
```
::::::::::::::::::::::::::::::::::::::::::::::::

## Visualization

"The `incidence2` object can be visualized using the `plot()` function from the base R package. 
The resulting graph is referred to as an epidemic curve, or epi-curve for short. The following code 
snippets generate epi-curves for the `dialy_incidence_data` and `weekly_incidence_data` incidence objects mentioned above."

```{r, message=FALSE, warning=FALSE}
# Load ggplot2 and tracetheme packages
library("ggplot2")
library("tracetheme")

# Plot daily incidence data
base::plot(dialy_incidence_data) +
  ggplot2::labs(
    x = "Time (in days)",
    y = "Dialy cases"
  ) +
  tracetheme::theme_trace()
``` 


```{r, message=FALSE, warning=FALSE}

# Plot weekly incidence data

base::plot(weekly_incidence_data) +
  ggplot2::labs(
    x = "Time (in weeks)",
    y = "weekly cases"
  ) +
  tracetheme::theme_trace()
``` 

## Curve of cumulative cases

The cumulative number of cases can calculated using `cumulate()` function from an `incidence2` object and visualize it, as in the below example.

```{r, message=FALSE, warning=FALSE}
cum_df <- incidence2::cumulate(dialy_incidence_data)
base::plot(cum_df) +
  ggplot2::labs(
    x = "Time (in days)",
    y = "weekly cases"
  ) +
  tracetheme::theme_trace()
```

Note that this function preserve grouping, i.e., if the the `incidence2` object contains groups, then it will accumlate the cases accrodingly. 

##  Peak estimation

One can estimate the peak --the time with highest number of recorded cases-- using `estimate_peak()` function from the {incidence2} package. 
This function employs a bootstrapping method to determine the peak time.

```{r, message=FALSE, warning=FALSE}
peak <- incidence2::estimate_peak(
  dialy_incidence_data,
  n = 100,
  alpha = 0.05,
  first_only = TRUE,
  progress = FALSE
)
print(peak)
```
This example demonstrates how to estimate the peak time using the `estimate_peak()` function at $95%$ 
confidence interval and using 100 bootstrap samples. 



::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 1: Can you do it?

 - Using suitable distributions for contacts, contact interval, infectious period, onset to hospitalized, and onset to 
 death, generate a simulated linelist data for  Marburg outbreak that has the probability of $0.5$ infection?
 - Aggregate the generated linelist and produce some epidemic curves?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints 

- Use `{simulist}` package to generate synthetic outbreak data
- Use `{incidence2}` package to aggregate case data based on a date event, and produce epidemic curves. 

::::::::::::::::::::::::::::::::::::::::::::::::
