---
title: 'Aggregate and visualize'
teaching: 20
exercises: 10
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to aggregate case data? 
- How to visualize aggregated data?
- What is distribution of cases in time, place, gender, age?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Simulate synthetic outbreak data
- Convert linelist data to incidence 
- Create epidemic curves from incidence data
::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

In an analytic pipeline, exploratory data analysis (EDA) is an important step before formal modelling. EDA helps 
determine relationships between variables and summarize their main characteristics often by means of data visualization. 

This episode focuses on EDA of outbreak data using a few essential R packages. 
A key aspect of EDA in epidemic analysis is identifying the relationship between time and the  observed epidemic outcome, such as confirmed cases, hospitalizations, deaths, and recoveries across different locations and demographic factors, including gender, age, and more.

Let's start by loading the package `{incidence2}` to aggregate linelist data by groups and visualize epicurves.
 We'll use `{simulist}` to simulate outbreak data,  and `{tracetheme}` for complementary figure formatting.
 We'll use the pipe `%>%` to connect some of their functions, including others from the packages `{dplyr}` and 
 `{ggplot2}`, so let's also call to the tidyverse package:

```{r,eval=TRUE,message=FALSE,warning=FALSE}
# Load packages
library(incidence2) # to aggregate and visualise
library(simulist) # to simulate linelist data
library(tracetheme) # for figure formatting
library(tidyverse) # for {dplyr} and {ggplot2} functions and the pipe %>%
```

::::::::::::::::::: checklist

### The double-colon

The double-colon `::` in R let you call a specific function from a package without loading the entire package into the current environment. 

For example, `dplyr::filter(data, condition)` uses `filter()` from the `{dplyr}` package.
This help us remember package functions and avoid namespace conflicts.

:::::::::::::::::::

 
## Synthetic outbreak data

To illustrate the process of conducting EDA on outbreak data, we will generate a line list 
for a hypothetical disease outbreak utilizing the `{simulist}` package. `{simulist}` generates simulation data for outbreak according to a given configuration. 
Its minimal configuration can generate a  linelist as shown in the below code chunk 

```{r, warning=FALSE, message=FALSE}
set.seed(1)
sim_data <- simulist::sim_linelist(outbreak_size = c(1000, 2000))
head(sim_data)
```

 This linelist dataset offers individual-level information about the outbreak. 

::::::::::::::::::: spoiler

## Additional Resources on Outbreak Data

 - This is the default configuration of `{simulist}`, should want to know more about its functionalities, 
check [middle](https://github.com/epiverse-trace/tutorials-middle/) and [late](https://epiverse-trace.github.io/tutorials-late/) tutorials.

 - You can find more information at the [`{outbreak}` documentation](https://outbreak-info.github.io/R-outbreak-info/)

:::::::::::::::::::



## Aggregating

Downstream analysis involves working with aggregated data rather than individual cases. This requires grouping linelist 
data in the form of incidence data. The [incidence2]((https://www.reconverse.org/incidence2/articles/incidence2.html){.external target="_blank"}) 
package offers an essential function, called `incidence`, for grouping case data, usually centered around dated events 
and/or other factors. The code chunk provided below demonstrates the creation of an `incidence2` object from the 
simulated  Ebola `linelist` data based on the  date of onset.

```{r, message=FALSE, warning=FALSE}
# create incidence object by aggregating case data  based on the date of onset
dialy_incidence_data <- incidence2::incidence(
  sim_data,
  date_index = "date_onset",
  interval = 1
)

# View the first incidence data for the first 5 days
dialy_incidence_data
```
Furthermore, with the `{incidence2}` package, you can specify the desired interval and categorize cases by one or 
more factors. Below is a code snippet demonstrating weekly cases grouped by the date of onset and gender.

```{r}
# Grouping data by week
weekly_incidence_data <- incidence2::incidence(
  sim_data,
  date_index = "date_onset",
  interval = 7,
  groups = c("sex", "case_type")
)

# View incidence data for the first 5 weeks
weekly_incidence_data
```

::::::::::::::::::::::::::::::::::::: callout
## Notes 
When cases are grouped by different factors, it's possible that these groups may have different date ranges in the 
resulting `incidence2` object. The `incidence2` package provides a function called `complete_dates()` to ensure that an
 incidence object has the same range of dates for each group. By default, missing counts will be filled with 0.

```{r, message=FALSE, warning=FALSE}
# Create incidence object
dialy_incidence_data_2 <- incidence2::incidence(
  sim_data,
  date_index = "date_onset",
  groups = "sex",
  interval = 1
)

# Complete missing dates in the incidence object
incidence2::complete_dates(
  x = dialy_incidence_data_2,
  expand = TRUE,
  fill = 0L, by = 1L,
  allow_POSIXct = FALSE
)
```
::::::::::::::::::::::::::::::::::::::::::::::::

## Visualization

"The `incidence2` object can be visualized using the `plot()` function from the base R package. 
The resulting graph is referred to as an epidemic curve, or epi-curve for short. The following code 
snippets generate epi-curves for the `dialy_incidence_data` and `weekly_incidence_data` incidence objects mentioned above."

```{r, message=FALSE, warning=FALSE}
# Plot daily incidence data
base::plot(dialy_incidence_data) +
  ggplot2::labs(
    x = "Time (in days)",
    y = "Dialy cases"
  ) +
  tracetheme::theme_trace()
``` 


```{r, message=FALSE, warning=FALSE}
# Plot weekly incidence data

base::plot(weekly_incidence_data) +
  ggplot2::labs(
    x = "Time (in weeks)",
    y = "weekly cases"
  ) +
  tracetheme::theme_trace()
``` 

## Curve of cumulative cases

The cumulative number of cases can be calculated using the `cumulate()` function from an `incidence2` object and visualized, as in the example below.

```{r, message=FALSE, warning=FALSE}
cum_df <- incidence2::cumulate(dialy_incidence_data)

base::plot(cum_df) +
  ggplot2::labs(
    x = "Time (in days)",
    y = "weekly cases"
  ) +
  tracetheme::theme_trace()
```

Note that this function preserves grouping, i.e., if the `incidence2` object contains groups, it will accumulate the cases accordingly. Give it a try with the `weekly_incidence_data` object!

##  Peak estimation

One can estimate the peak --the time with the highest number of recorded cases-- using the `estimate_peak()` function from the {incidence2} package. 
This function employs a bootstrapping method to determine the peak time.

```{r, message=FALSE, warning=FALSE}
peak <- incidence2::estimate_peak(
  dialy_incidence_data,
  n = 100,
  alpha = 0.05,
  first_only = TRUE,
  progress = FALSE
)

peak
```
This example demonstrates how to estimate the peak time using the `estimate_peak()` function at $95%$ 
confidence interval and using 100 bootstrap samples. 

## Visulaziantion with ggplot2


`{incidence2}` produces basic plots for epicurves, but additional work is required to create well-annotated graphs. However, using the `{ggplot2}` package, you can generate more sophisticated and better-annotated epicurves.
`{ggplot2}` is a comprehensive package with many functionalities. However, we will focus on three key elements for producing epicurves: histogram plots, scaling date axes and their labels, and general plot theme annotation.
The example below demonstrates how to configure these three elements for a simple `{incidence2}` object.

```{r, message=FALSE, warning=FALSE}
breaks <- seq.Date(
  from = min(as.Date(dialy_incidence_data$date_index,
    na.rm = TRUE
  )),
  to = as.Date(max(dialy_incidence_data$date_index,
    na.rm = TRUE
  )),
  by = 1
)

ggplot2::ggplot(data = dialy_incidence_data) +
  geom_histogram(
    mapping = aes(
      x = as.Date(date_index),
      y = count
    ),
    stat = "identity",
    color = "blue",
    width = 1
  ) +
  theme_minimal() + # simple theme
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.caption = element_text(face = "italic", hjust = 0),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45)
  ) +
  labs(
    x = "Date", # x-label
    y = "Number of cases", # y-label,
    title = "Daily outbreak cases", # title
    subtitle = "subtitle", # subtitle
    caption = "informative caption"
  ) +
  scale_x_date(
    breaks = breaks,
    label = scales::label_date_short()
  )
```

Use the `group` option in the mapping function to visualize an epicurve with different groups. If there is more than one grouping factor, use the `facet_wrap()` option, as demonstrated in the example below:

```{r, message=FALSE, warning=FALSE}
ggplot2::ggplot(data = dialy_incidence_data_2) +
  geom_histogram(
    mapping = aes(
      x = as.Date(date_index),
      y = count,
      group = sex,
      fill = sex
    ),
    stat = "identity"
  ) +
  theme_minimal() + # simple theme
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.caption = element_text(face = "italic", hjust = 0),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45)
  ) +
  labs(
    x = "Date", # x-label
    y = "Number of cases", # y-label,
    title = "Daily outbreak cases", # title
    subtitle = "subtitle", # subtitle
    caption = "informative caption"
  ) +
  facet_wrap(~sex) +
  scale_x_date(
    breaks = breaks,
    label = scales::label_date_short()
  )
```


::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 1: Can you do it?

 - Using suitable distributions for contacts, contact interval, infectious period, onset to hospitalized, and onset to 
 death, generate a simulated linelist data for  Marburg outbreak that has the probability of $0.5$ infection?
 - Aggregate the generated linelist and produce some epidemic curves?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints 

- Use `{simulist}` package to generate synthetic outbreak data
- Use `{incidence2}` package to aggregate case data based on a date event, and produce epidemic curves. 

::::::::::::::::::::::::::::::::::::::::::::::::
