---
title: 'Aggregate and visulaize'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to aggregate case data? 
- How to visualize aggregated data?
- What is distribution of cases in time, place, gender, age?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Convert case data to incidence 
- Create epidemic curves from incidence data
- Estimate the growth rate from incidence curves
- Create quick descriptive and comparison tables

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

A comprehensive description of data is pivotal for conducting insightful explanatory and exploratory analyses. This episode focuses on describing and visualizing epidemic data. The examples are built around the **Covid-19 case data from England** dataset that contained in the [outbreaks](http://www.reconverse.org/outbreaks/) package. The first setp is to  read this dataset, and we recommend utilizing the [eadr](../links.md#readr) package for this purpose (or employing alternative methods as outlined in the [Read case data](../episodes/read-cases.Rmd) episode.

```{r, warning=FALSE, message=FALSE}
requireNamespace("outbreaks")
covid19_eng_case_data <- outbreaks::covid19_england_nhscalls_2020
utils::head(covid19_eng_case_data, 5)
```

## Incidence data

Downstream analysis involves working with aggregated data rather than individual cases. This requires aggregating case data and creating incidence data. The [incidence2]((https://www.reconverse.org/incidence2/articles/incidence2.html){.external target="_blank"}) package offers an essential function, called `incidence`, for grouping case data, usually centered around dated events and/or other factors. The code chunk provided below demonstrates the creation of an `incidence2` object from the `covid19_eng_case_data` based on the date of samples.

```{r, message=FALSE, warning=FALSE}
requireNamespace("incidence2")
covid19_eng_incidence_data <- incidence2::incidence(
  covid19_eng_case_data,
  date_index = "date"
)
utils::head(covid19_eng_incidence_data, 5)
```

The `incidence2` object can be visualized using the `plot()` function from base R package. 

```{r, message=FALSE, warning=FALSE}
base::plot(covid19_eng_incidence_data)
```

Moreover, `{incidence2}` has other functions that allow for aggregating case data based on a dated event and other factors such as the individual gender, the sampling location, etc. In the example below, we calculate weekly counts of Covid-19 cases in England grouping them by `sex` type.

```{r, message=FALSE, warning=FALSE}
weekly_covid19_eng_incidence <- incidence2::incidence(
  covid19_eng_case_data,
  date_index = "date",
  interval = "week",
  groups = "sex"
)
base::plot(weekly_covid19_eng_incidence, angle = 45)
```



::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 1: Can you do it?

 - Using the above `covid91_eng_case_data`  dataset, produce monthly epi-curves for Covid-19 cases in England based on regional places in England?

:::::::::::::::::::::::: solution 

 
```r
monthly_covid19_eng_incidence <- incidence2::incidence(
  covid19_eng_case_data,
  date_index = "date",
  interval = "month",
  groups = "sex"
)
base::plot(monthly_covid19_eng_incidence, angle = 45)
```

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::: keypoints 

- Use `{incidence2}` to aggregate case data based on a date event.  
::::::::::::::::::::::::::::::::::::::::::::::::

