---
title: 'Aggregate and visulaize'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to aggregate case data? 
- How to visualize aggregated data?
- What is distribution of cases in time, place, gender, age?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Convert case data to incidence 
- Create epidemic curves from incidence data
- Estimate the growth rate from incidence curves
- Create quick descriptive and comparison tables

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction
In an analytic pipeline, exploratory data analysis (EDA) is an important step before formal modelling. EDA helps 
determine relationships between variables and summarize their main characteristics often by means of data visualization. 

This episode focuses on  EDA of outbreaks and epidemic data, and how to achieved that using a couples of handy `R` 
packages. A key observation in EDA of epidemic analysis is capturing the relationship between time and the number of 
reported cases, spanning various categories (confirmed, hospitalized, deaths, and recoveries), locations, and other 
demographic factors such as gender, age, etc.  
 
## Synthetic outbreak 
To illustrate the process of conducting EDA on outbreak data, we will generate a line list 
for a hypothetical Ebola outbreak utilizing the `{simulist}` package. This line list dataset offers individual-level 
information about the outbreak. For our simulation, we will assume that the dynamics of this outbreak are influenced by 
several factors: the contact distribution (average number of contacts for an infected case), distribution of contact 
intervals (time period between contacts), and the delay distributions of onset to hospitalization and onset to death. 
These latter distributions can be sourced from literature and are conveniently available in the `{epiparameter}` 
package, see the below code chunk.

```{r, warning=FALSE, message=FALSE}
requireNamespace("simulist", quietly = TRUE)
requireNamespace("epiparameter", quietly = TRUE)
contact_dist <- epiparameter::epidist(
  disease = "Ebola",
  epi_dist = "contact distribution",
  prob_distribution = "pois",
  prob_distribution_params = c(mean = 4)
)
cont_interval <- epiparameter::epidist(
  disease = "Ebola",
  epi_dist = "contact interval",
  prob_distribution = "gamma",
  prob_distribution_params = c(shape = 1, scale = 1)
)
# Define onset to hospitalized
onset_to_hosp <- contact_dist <- epiparameter::epidist(
  disease = "Ebola",
  epi_dist = "onset to hospitalisatio",
  prob_distribution = "pois",
  prob_distribution_params = c(mean = 14)
)

# get onset to death from {epiparameter} database
onset_to_death <- epiparameter::epidist_db(
  disease = "Ebola",
  epi_dist = "onset to death",
  single_epidist = TRUE
)
infect_period <- epiparameter::epidist(
  disease = "Ebola",
  epi_dist = "Infectious period",
  prob_distribution = "gamma",
  prob_distribution_params = c(shape = 1, scale = 1)
)
```
Additionally, we assume that the outbreak started at the beginning of 2023, is highly contagious with a probability of 
infection of 80%, and its minimum and maximum sizes are 1000 and 10,000, respectively. Combining these assumptions with 
the mentioned distributions, the code chunk below generates a simulated line list:
```{r, warning=FALSE, message=FALSE}
base::set.seed(1)
linelist <- simulist::sim_linelist(
  contact_dist,
  infect_period,
  prob_infect = 0.8,
  onset_to_hosp,
  onset_to_death,
  hosp_risk = 0.2,
  hosp_death_risk = 0.5,
  non_hosp_death_risk = 0.05,
  outbreak_start_date = as.Date("2023-01-01"),
  add_names = TRUE,
  add_ct = TRUE,
  outbreak_size = c(1000, 10000),
  population_age = c(1, 90),
  case_type_probs = c(suspected = 0.2, probable = 0.1, confirmed = 0.7),
  config = simulist::create_config()
)
utils::head(linelist)
```

## Old code
```{r, warning=FALSE, message=FALSE}
requireNamespace("outbreaks", quietly = TRUE)
covid19_eng_case_data <- outbreaks::covid19_england_nhscalls_2020
utils::head(covid19_eng_case_data, 5)
```

## Incidence data

Downstream analysis involves working with aggregated data rather than individual cases. This requires aggregating case data in the form of incidence data. The [incidence2]((https://www.reconverse.org/incidence2/articles/incidence2.html){.external target="_blank"}) package offers an essential function, called `incidence`, for grouping case data, usually centered around dated events and/or other factors. The code chunk provided below demonstrates the creation of an `incidence2` object from the `covid19_eng_case_data` based on the sampling date.

```{r, message=FALSE, warning=FALSE}
requireNamespace("incidence2", quietly = TRUE)
covid19_eng_incidence_data <- incidence2::incidence(
  covid19_eng_case_data,
  date_index = "date"
)
utils::head(covid19_eng_incidence_data, 5)
```

The `incidence2` object can be visualized using the `plot()` function from base R package. 

```{r, message=FALSE, warning=FALSE}
base::plot(covid19_eng_incidence_data)
```

Moreover, `{incidence2}` has other functions that allow for aggregating case data based on a dated event and other factors such as the individual gender, the sampling location, etc. In the example below, we calculate weekly counts of Covid-19 cases in England grouping them by `sex` type.

```{r, message=FALSE, warning=FALSE}
weekly_covid19_eng_incidence <- incidence2::incidence(
  covid19_eng_case_data,
  date_index = "date",
  interval = "week",
  groups = "sex"
)
base::plot(weekly_covid19_eng_incidence, angle = 45)
```



::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 1: Can you do it?

 - Using the above `covid91_eng_case_data`  dataset, produce monthly epi-curves for Covid-19 cases in England based on regional places in England?

:::::::::::::::::::::::: solution 

 
```{r}
monthly_covid19_eng_incidence <- incidence2::incidence(
  covid19_eng_case_data,
  date_index = "date",
  interval = "month",
  groups = "sex"
)
base::plot(monthly_covid19_eng_incidence, angle = 45)
```

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::: keypoints 

- Use `{incidence2}` to aggregate case data based on a date event.  
::::::::::::::::::::::::::::::::::::::::::::::::

