---
title: 'Aggregate and visualize'
teaching: 20
exercises: 10
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to aggregate and summarise case data? 
- How to visualize aggregated data?
- What is distribution of cases in time, place, gender, age?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Simulate synthetic outbreak data
- Convert indivdual linelist data to incidence over time
- Create epidemic curves from incidence data
::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

In an analytic pipeline, exploratory data analysis (EDA) is an important step before formal modelling. EDA helps 
determine relationships between variables and summarize their main characteristics, often by means of data visualization. 

This episode focuses on EDA of outbreak data using R packages. 
A key aspect of EDA in epidemic analysis is 'person, place and time'. It is useful to identify how observed events - such as confirmed cases, hospitalizations, deaths, and recoveries - change over time, and how these vary across different locations and demographic factors, including gender, age, and more.

Let's start by loading the package `{incidence2}` to aggregate linelist data according to specific characteristics, and visualize the resulting epidemic curves (epicurves) that plot the number of new events (i.e. incidence) over time.
 We'll use `{simulist}` to simulate some outbreak data to analyse,  and `{tracetheme}` for figure formatting.
 We'll use the pipe `%>%` to connect some of their functions, including others from the packages `{dplyr}` and 
 `{ggplot2}`, so let's also call to the tidyverse package:

```{r,eval=TRUE,message=FALSE,warning=FALSE}
# Load packages
library(incidence2) # For aggregating and visualising
library(simulist) # For simulating linelist data
library(tracetheme) # For formatting figures
library(tidyverse) # For {dplyr} and {ggplot2} functions and the pipe %>%
```

::::::::::::::::::: checklist

### The double-colon

The double-colon `::` in R lets you call a specific function from a package without loading the entire package into the current environment. 

For example, `dplyr::filter(data, condition)` uses `filter()` from the `{dplyr}` package.
This help us remember package functions and avoid namespace conflicts.

:::::::::::::::::::

 
## Synthetic outbreak data

To illustrate the process of conducting EDA on outbreak data, we will generate a line list 
for a hypothetical disease outbreak utilizing the `{simulist}` package. `{simulist}` generates simulation data for outbreak according to a given configuration. 
Its minimal configuration can generate a linelist, as shown in the below code chunk :

```{r}
# Simulate linelist data for an outbreak with size between 1000 and 1500
set.seed(1) # Set seed for reproducibility
sim_data <- simulist::sim_linelist(outbreak_size = c(1000, 1500)) %>%
  dplyr::as_tibble() # for a simple data frame output

# Display the simulated dataset
sim_data
```

This linelist dataset has entries on individual-level simulated events during the outbreak.

::::::::::::::::::: spoiler

## Additional Resources on Outbreak Data

The above is the default configuration of `{simulist}`, so includes a number of assumptions about the transmissibility and severity of the pathogen. If you want to know more about `sim_linelist()` and other functionalities
check the [documentation website](https://epiverse-trace.github.io/simulist/).

You can also find data sets from real emergencies from the past at the [`{outbreaks}` R package](https://www.reconverse.org/outbreaks/).

:::::::::::::::::::



## Aggregating

Often we want to analyse and visualise the number of events that occur on a particular day or week, rather than focusing on individual cases. This requires grouping linelist 
data into incidence data. The [incidence2]((https://www.reconverse.org/incidence2/articles/incidence2.html){.external target="_blank"}) 
package offers a useful function called `incidence2::incidence()` for grouping case data, usually based around dated events 
and/or other characteristics. The code chunk provided below demonstrates the creation of an `<incidence2>` class object from the 
simulated  Ebola `linelist` data based on the date of onset.

```{r}
# Create an incidence object by aggregating case data based on the date of onset
daily_incidence <- incidence2::incidence(
  sim_data,
  date_index = "date_onset",
  interval = "day" # Aggregate by daily intervals
)

# View the incidence data
daily_incidence
```
With the `{incidence2}` package, you can specify the desired interval (e.g. day, week) and categorize cases by one or 
more factors. Below is a code snippet demonstrating weekly cases grouped by the date of onset, sex, and type of case.

```{r}
# Group incidence data by week, accounting for sex and case type
weekly_incidence <- incidence2::incidence(
  sim_data,
  date_index = "date_onset",
  interval = "week", # Aggregate by weekly intervals
  groups = c("sex", "case_type") # Group by sex and case type
)

# View the incidence data
weekly_incidence
```

::::::::::::::::::::::::::::::::::::: callout
## Dates Completion  
When cases are grouped by different factors, it's possible that the events involving these groups may have different date ranges in the 
resulting `incidence2` object. The `incidence2` package provides a function called `complete_dates()` to ensure that an
 incidence object has the same range of dates for each group. By default, missing counts for a particular group will be filled with 0 for that date.
 
This functionality is also available as an argument within `incidence2::incidence()` adding `complete_dates = TRUE`.

```{r}
# Create an incidence object grouped by sex, aggregating daily
daily_incidence_2 <- incidence2::incidence(
  sim_data,
  date_index = "date_onset",
  groups = "sex",
  interval = "day", # Aggregate by daily intervals
  complete_dates = TRUE # Complete missing dates in the incidence object
)
```

```{r,echo=FALSE,eval=FALSE}
daily_incidence_2_complete <- incidence2::complete_dates(
  x = daily_incidence_2,
  expand = TRUE, # Expand to fill in missing dates
  fill = 0L,     # Fill missing values with 0
  by = 1L,       # Fill by daily intervals
  allow_POSIXct = FALSE # Ensure that dates are not in POSIXct format
)
```


::::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 1: Can you do it?
 - **Task**: Aggregate `sim_data` linelist based on admission date and case outcome in __biweekly__
  intervals, and save the results in an object called `biweekly_incidence`.

::::::::::::::::::::::::::::::::::::::::::::::::

## Visualization

The `incidence2` object can be visualized using the `plot()` function from the base R package. 
The resulting graph is referred to as an epidemic curve, or epi-curve for short. The following code 
snippets generate epi-curves for the `daily_incidence` and `weekly_incidence` incidence objects mentioned above.

```{r}
# Plot daily incidence data
base::plot(daily_incidence) +
  ggplot2::labs(
    x = "Time (in days)", # x-axis label
    y = "Dialy cases" # y-axis label
  ) +
  tracetheme::theme_trace() # Apply the custom trace theme
``` 


```{r}
# Plot weekly incidence data
base::plot(weekly_incidence) +
  ggplot2::labs(
    x = "Time (in weeks)", # x-axis label
    y = "weekly cases" # y-axis label
  ) +
  tracetheme::theme_trace() # Apply the custom trace theme
``` 

:::::::::::::::::::::::: callout

#### easy aesthetics

We invite you to skim the `{incidence2}` package ["Get started" vignette](https://www.reconverse.org/incidence2/articles/incidence2.html). Find how you can use arguments within `plot()` to provide aesthetics to your incidence2 class objects!

```{r}
base::plot(weekly_incidence, fill = "sex")
```

Some of them include `show_cases = TRUE`, `angle = 45`, and `n_breaks = 5`. Give them a try!

::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 2: Can you do it?
 - **Task**: Visualize `biweekly_incidence` object.

::::::::::::::::::::::::::::::::::::::::::::::::

## Curve of cumulative cases

The cumulative number of cases can be calculated using the `cumulate()` function from an `incidence2` object and visualized, as in the example below.

```{r}
# Calculate cumulative incidence
cum_df <- incidence2::cumulate(daily_incidence)

# Plot cumulative incidence data using ggplot2
base::plot(cum_df) +
  ggplot2::labs(
    x = "Time (in days)", # x-axis label
    y = "weekly cases" # y-axis label
  ) +
  tracetheme::theme_trace() # Apply the custom trace theme
```

Note that this function preserves grouping, i.e., if the `incidence2` object contains groups, it will accumulate the cases accordingly.


::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 3: Can you do it?
 - **Task**: Visulaize the cumulatie cases from `biweekly_incidence` object.

::::::::::::::::::::::::::::::::::::::::::::::::

##  Peak estimation

One can estimate the peak --the time with the highest number of recorded cases-- using the `estimate_peak()` function from the {incidence2} package. 
This function employs a bootstrapping method to determine the peak time.

```{r}
# Estimate the peak of the daily incidence data
peak <- incidence2::estimate_peak(
  daily_incidence,
  n = 100,         # Number of simulations for the peak estimation
  alpha = 0.05,    # Significance level for the confidence interval
  first_only = TRUE, # Return only the first peak found
  progress = FALSE  # Disable progress messages
)

# Display the estimated peak
print(peak)
```
This example demonstrates how to estimate the peak time using the `estimate_peak()` function at $95%$ 
confidence interval and using 100 bootstrap samples. 

::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 4: Can you do it?
 - **Task**: Estimate the peak time from `biweekly_incidence` object.

::::::::::::::::::::::::::::::::::::::::::::::::


## Visualization with ggplot2


`{incidence2}` produces basic plots for epicurves, but additional work is required to create well-annotated graphs. However, using the `{ggplot2}` package, you can generate more sophisticated and better-annotated epicurves.
`{ggplot2}` is a comprehensive package with many functionalities. However, we will focus on three key elements for producing epicurves: histogram plots, scaling date axes and their labels, and general plot theme annotation.
The example below demonstrates how to configure these three elements for a simple `{incidence2}` object.

```{r}
# Define date breaks for the x-axis
breaks <- seq.Date(
  from = min(as.Date(daily_incidence$date_index, na.rm = TRUE)),
  to = max(as.Date(daily_incidence$date_index, na.rm = TRUE)),
  by = 20 # every 20 days
)

# Create the plot
ggplot2::ggplot(data = daily_incidence) +
  geom_histogram(
    mapping = aes(
      x = as.Date(date_index),
      y = count
    ),
    stat = "identity",
    color = "blue", # bar border color
    fill = "lightblue", # bar fill color
    width = 1 # bar width
  ) +
  theme_minimal() + # apply a minimal theme for clean visuals
  theme(
    plot.title = element_text(face = "bold",
                              hjust = 0.5), # center and bold title
    plot.subtitle = element_text(hjust = 0.5), # center subtitle
    plot.caption = element_text(face = "italic",
                                hjust = 0), # italicized caption
    axis.title = element_text(face = "bold"), # bold axis titles
    axis.text.x = element_text(angle = 45, vjust = 0.5) # rotated x-axis text
  ) +
  labs(
    x = "Date", # x-axis label
    y = "Number of cases", # y-axis label
    title = "Daily Outbreak Cases", # plot title
    subtitle = "Epidemiological Data for the Outbreak", # plot subtitle
    caption = "Data Source: Simulated Data" # plot caption
  ) +
  scale_x_date(
    breaks = breaks, # set custom breaks on the x-axis
    labels = scales::label_date_short() # shortened date labels
  )
```

Use the `group` option in the mapping function to visualize an epicurve with different groups. If there is more than one grouping factor, use the `facet_wrap()` option, as demonstrated in the example below:

```{r}
# Plot daily incidence by sex with facets
ggplot2::ggplot(data = daily_incidence_2) +
  geom_histogram(
    mapping = aes(
      x = as.Date(date_index),
      y = count,
      group = sex,
      fill = sex
    ),
    stat = "identity"
  ) +
  theme_minimal() + # apply minimal theme
  theme(
    plot.title = element_text(face = "bold",
                              hjust = 0.5), # bold and center the title
    plot.subtitle = element_text(hjust = 0.5), # center the subtitle
    plot.caption = element_text(face = "italic", hjust = 0), # italic caption
    axis.title = element_text(face = "bold"), # bold axis labels
    axis.text.x = element_text(angle = 45,
                               vjust = 0.5) # rotate x-axis text for readability
  ) +
  labs(
    x = "Date", # x-axis label
    y = "Number of cases", # y-axis label
    title = "Daily Outbreak Cases by Sex", # plot title
    subtitle = "Incidence of Cases Grouped by Sex", # plot subtitle
    caption = "Data Source: Simulated Data" # caption for additional context
  ) +
  facet_wrap(~sex) + # create separate panels by sex
  scale_x_date(
    breaks = breaks, # set custom date breaks
    labels = scales::label_date_short() # short date format for x-axis labels
  ) +
  scale_fill_manual(values = c("lightblue",
                               "lightpink")) # custom fill colors for sex
```


::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 5: Can you do it?
 - **Task**: Produce an annotated figure for biweekly_incidence using `{ggplot2}` package.

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints 

- Use `{simulist}` package to generate synthetic outbreak data
- Use `{incidence2}` package to aggregate case data based on a date event, and produce epidemic curves. 
- Use `{ggplot2}` package to produce better annotated epicurves. 

::::::::::::::::::::::::::::::::::::::::::::::::
