---
title: 'Read case data'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- how many different data formats can I read? 
- is it possible to import data from database and health APIs? 
::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Explain how read/import outbreak data from different sources into `R` 
environment for analysis.
::::::::::::::::::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites

This episode requires you to be familiar with:

**Data science** : Basic programming with R.
:::::::::::::::::::::::::::::::::

# Introduction

The initial step in outbreak analysis involves importing the target dataset into the `R` environment from various sources. Outbreak data is typically stored in files of diverse formats,  relational database management systems (RDBMS) servers, and health information system (HIS) application program interfaces (APIs) such as [REDCap](https://www.project-redcap.org/) and [DHIS2](https://dhis2.org/). The latter two options are particularly well-suited for storing institutional data. This episode will elucidate the process of reading cases from these sources.

## Reading from files

Several packages are available for importing outbreak data stored in individual files into `R`. These include [rio](http://gesistsa.github.io/rio/), [readr](https://readr.tidyverse.org/) from the `tidyverse`, [io](https://bitbucket.org/djhshih/io/src/master/), and [ImportExport](https://cran.r-project.org/web/packages/ImportExport/index.html). Together, these packages offer methods to read single or multiple files in a wide range of formats.

The below example shows how to import a `csv` file into `R` environment using `rio` package.
```{r,warning=FALSE,message=FALSE}
library("rio")
case_data = rio::import("./data/ebola_cases.csv", format = "csv")
```
Similarly, you can import files of other formats such as "tsv", "xlsx", ...etc.

::::::::::::::::::::::::::::::::: challenge

###  Reading compressed data 
Take 1 minute:

- Is it possible to read compressed data in `R`?

::::::::::::::::: hint

You can check the supported file formats in the `{rio}` package as follows:
```{r, eval=FALSE}
library("rio")
rio::install_formats()
```

::::::::::::::::::::::

::::::::::::::::: solution

```{r,eval=FALSE}
library("rio")
rio::import("/path_name/file_name.zip")
```

::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::


## Reading from databases

The [DBI](https://dbi.r-dbi.org/) package serves as a versatile interface for interacting with database management systems (DBMS) across different back-ends or servers. It offers a uniform method for accessing and retrieving data from various database systems.


The following code chunk demonstrates how to create a temporary SQLite database in memory, store the `case_data` dataframe as a table within it, and subsequently read from it:

```{r,warning=FALSE,message=FALSE}
library(DBI)
library(RSQLite)
# Create a temporary SQLite database in memory
db_con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
# Store the 'case_data' dataframe as a table named 'cases' in the SQLite database
DBI::dbWriteTable(db_con, "cases", case_data)
# Read data from the 'cases' table
result <- DBI::dbReadTable(db_con, "cases")
# Close the database connection
DBI::dbDisconnect(db_con)
# View the result
base::print(utils::head(result))
```

This code first establishes a connection to an SQLite database created in memory using `dbConnect` function. Then, it writes the `case_data` dataframe into a table named 'cases' within the database using `dbWriteTable` function. Subsequently, it reads the data from the 'cases' table using `dbReadTable` function. Finally, it closes the database connection with `dbDisconnect` function.

More examples about SQL databases and R can be found [here](https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html).

## Reading from HIS APIs

Coming soon with {readepi}

::::::::::::::::::::::::::::::::::::: keypoints 
- Use `{rio}, {io}, {readr}` and `{ImportExport}` to read data from individual files.
- Use `{DBI}` to read data from databases.
- Use `{readepi}` to read data form HIS APIs.
::::::::::::::::::::::::::::::::::::::::::::::::

